% =====================================================================
%  §1 - Introduction
%
% =====================================================================

\section{Introduction}
\label{sec:intro}

Scientific and clinical decision-making depends on evidence from
heterogeneous sources, especially the primary literature. Today,
this evidence is largely consumed by humans and encoded indirectly
into annotation databases and local rules.

Existing resources such as HL7 Fast Healthcare Interoperability
Resources (FHIR) Evidence and EvidenceVariable~\cite{fhir_evidence},
the Evidence and Conclusion Ontology (ECO)~\cite{eco2019}, the
Scientific Evidence and Provenance Information Ontology
(SEPIO)~\cite{sepio}, and the Global Alliance for Genomics and
Health (GA4GH) Genomic Knowledge Standards (GKS)~\cite{ga4gh_va}
advance interoperability in clinical research. However, they do not
comprehensively address the evidence types and workflows that
originate in the basic sciences. Existing standards focus mainly on
clinical trials, high-level assertions, and variant representations,
and provide limited support for the fine-grained, domain-specific
characterisation of basic and pre-clinical scientific evidence needed
for automated reasoning and AI-ready infrastructures.

Each scientific domain demands a tailored approach for structuring
and communicating evidence, and this need is particularly prominent
in clinical genomics, which is the focus of this paper. Genetic
evidence is produced under diverse contexts, from experimental model
systems to large-scale population studies, that are not fully
supported by current standards. To properly use evidence derived from
scientific literature in clinical workflows, two principal challenges
require standardisation: \emph{Relevance and Specificity}, that is,
how precisely a genetic finding maps to an individual patient,
disease context, or biological hypothesis; and \emph{Confidence},
that is, whether a given evidence assertion can be robustly and
safely used, ensuring defensible compliance, interpretability, and
reproducibility.

A well-designed schema for genetic evidence helps make literature
review a \emph{semantic-parsing} problem: mapping natural-language
claims to a shared, source-anchored representation. Large language
models are increasingly capable of this kind of parsing, but
dependable automation still requires careful ontological grounding of
the target language. The present work offers one such increment,
building on decades of work by the biomedical-ontology community and
aiming toward literature review that is materially assisted by
automated parsing.

\paragraph{Contributions.} This paper makes the following contributions:
\begin{itemize}
    \item A domain-analysis-driven semantic model for genetic
    evidence, with core classes (\ScientificEvidence,
    \EvidenceVariable, \EvidenceAssertion) and genetic
    specialisations, plus a compact dimensional vocabulary
    formalised using the W3C Shapes Constraint Language (SHACL)%
    ~\cite{w3c-shacl}.
    \item A worked mapping of the model to six genetic-evidence papers
    spanning human, population, model-organism, and polygenic-score
    genetics, with annotations and extraction tooling.
    \footnote{\url{https://github.com/ForomePlatform/genetic-evidence-model}}
    \item A human--AI collaborative annotation workflow.
    \item An alignment sketch between the model and existing standards
    (FHIR Evidence, ECO, SEPIO, GA4GH GKS, ACMG/AMP, IAO, OBI,
    PROV-O, ClinGen, MONDO, HPO), intended as a compact pointer
    rather than an exhaustive mapping.
\end{itemize}

The remainder of the paper reviews related standards
(\Cref{sec:background}), defines the semantic model
(\Cref{sec:model}), and illustrates it on two worked examples
(\Cref{sec:examples}). We then describe the annotation workflow and
pilot statistics (\Cref{sec:workflow}), discuss findings, candidate
extensions, limitations, and future work (\Cref{sec:discussion}), and
conclude (\Cref{sec:conclusion}).
