% =====================================================================
%  §2 — Background and Related Work
%
%  Six standards families: FHIR Evidence, ECO, SEPIO, GA4GH GKS,
%  ACMG-AMP, IAO/OBI/PROV-O. Each in a compact paragraph. Closes
%  with a paragraph naming the gap this paper addresses.
%
%  Revised 2026-04-21: standards-mapping table dropped for space
%  (paper was at 18pp vs 12pp cap); per-family paragraphs tightened
%  and acronyms expanded consistently on first mention.
% =====================================================================

\section{Background and Related Work}
\label{sec:background}

The representation of scientific and clinical evidence has been the
subject of sustained standards work in at least three communities:
clinical informatics, biomedical ontology, and genomic knowledge
representation, each with its own emphasis. Our schema sits at the
intersection of these. Like clinical standards, it is oriented toward
supporting downstream decision-making; like ontology work, it is
grounded in formal constraints and explicit provenance; like genomic
knowledge representation, it treats target resolution (variant, gene,
region, whole-genome aggregate, and so on) as a first-class concern,
while relying on GA4GH standards for sequence identifiers. This
section positions the proposed schema among the six standards families
we have engaged with and closes by spelling out the specific coverage
gap that motivated the work. Detailed term-level alignment with each
family is scoped as future work (\cref{sec:disc-future}).

\paragraph{FHIR Evidence Resources.}
The HL7 FHIR Evidence, EvidenceVariable, and EvidenceReport
resources~\cite{fhir_evidence} provide structured representations of
clinical-trial-style evidence, with explicit fields for population,
intervention, comparator, outcome, and statistical measures.
FHIR Evidence was designed around systematic-review and
guideline-development workflows and assumes that evidence concerns an
intervention tested in a population. This assumption does not fit
basic-science and pre-clinical genetic evidence cleanly: a mouse
knockout study does not have an ``intervention'' in the clinical-trial
sense, and a coding-variant case--control study has a target (the
variant) that is not the same kind of object as a drug or a procedure.
Our schema shares FHIR Evidence's commitment to structured,
dimensional description but replaces this clinical-trial skeleton with
a vocabulary appropriate to genetic claims.

\paragraph{Evidence Codes: ECO and SEPIO.}
The Evidence and Conclusion Ontology (ECO)~\cite{eco2019} provides a
large hierarchical vocabulary of evidence types, including inference
methods, assay types, and source-document categories, organised within
the Open Biological and Biomedical Ontologies (OBO) framework and
widely used in Gene Ontology annotation. The Scientific Evidence and
Provenance Information Ontology (SEPIO)~\cite{sepio} complements ECO
with machine-readable representations of statements, qualifiers, and
provenance relationships, supporting structured evidence chains. Our
Method and Measurement Target dimensions overlap substantially with
areas that ECO and SEPIO already populate, and our long-term goal is
to ground those dimensions in ECO and SEPIO identifiers rather than
maintain a separate vocabulary. The contribution of the present schema
relative to ECO and SEPIO is therefore orthogonal: we provide the
conditional-activation structure and the paper-level decomposition into
\GeneticEvidence{} items, both of which are independent of the
particular evidence-code vocabulary used to populate the Method
dimension.

\paragraph{GA4GH Genomic Knowledge Standards.}
The most natural point of comparison for our schema within the
GA4GH Genomic Knowledge Standards effort is the Variant Annotation
(VA) Specification~\cite{ga4gh_va}. VA is a broad information model
for structured statements about variants, organised around a
\texttt{Statement} class whose instances carry a proposition,
direction of support, strength, classification, a list of evidence
lines, and contribution and method metadata. In scope and
orientation VA is information-centric: its top-level abstraction is
an \texttt{InformationEntity}, and it represents what is known about
a variant as structured information rather than the process by which
primary-literature claims become evidence for downstream reasoning.
Its design does not engage explicitly with FHIR Evidence or kindred
evidence standards, and it is not primarily intended for parsing a
publication and assessing the credibility of what it reports. Our
schema is narrower and complementary: it focuses on evidence drawn
from primary-literature publications, with the \emph{Relevance and
Specificity} and \emph{Confidence} challenges of \cref{sec:intro} as
concrete targets and an eventual semantic-parsing workflow over the
literature as the longer-horizon motivation. A VA \texttt{Statement}
can be viewed as a plausible downstream aggregate of the
per-publication evidence items our schema records, rather than a
substitute for them. The Variation Representation Specification
(VRS)~\cite{ga4gh_vrs}, a separate product of the same effort,
addresses canonical variant identity across coordinate systems and is
orthogonal to the concerns of the present work.

\paragraph{ACMG/AMP variant interpretation.}
The American College of Medical Genetics and Genomics /
Association for Molecular Pathology (ACMG/AMP)
guidelines~\cite{acmg2015} specify a rule-based framework for
variant classification from weighted evidence codes and
criteria. Our schema is intended to feed such frameworks rather
than replace them: its per-publication evidence items are the
kind of input an ACMG/AMP-compatible classifier would consume.
A term-level mapping from our Method and Credibility dimensions
to the ACMG/AMP evidence codes is scoped as future work.

\paragraph{Upper ontologies: IAO, OBI, PROV-O.}
The Information Artefact Ontology (IAO)~\cite{iao2015} and the
Ontology for Biomedical Investigations (OBI)~\cite{obi2016} provide
upper-level scaffolding for many OBO-compliant biomedical ontologies:
IAO supplies classes such as \emph{information-content-entity} and
\emph{textual-entity}, and OBI supplies the assay and protocol
hierarchies relevant to our Method dimension. The W3C PROV Ontology
(PROV-O)~\cite{prov_o} provides standard provenance relationships
(\emph{wasDerivedFrom}, \emph{wasGeneratedBy}, etc.) that a future
Web Ontology Language (OWL) representation of our schema would use for
source-span and curator-provenance fields. Formal alignment of our
top-level classes with IAO, OBI, and PROV-O is scoped as future work;
the current SHACL representation is intentionally lightweight, and we
expect to replace it with an OWL/LinkML rendering.

\paragraph{Related annotation efforts.}
The Clinical Genome Resource (ClinGen) evidence curation
framework~\cite{clingen2018} provides a structured workflow for
gene--disease validity and variant pathogenicity assessments with
explicit evidence rubrics; our schema shares its commitment to
source-anchored claims and explicit categorisation but operates at a
lower level by representing the individual claims that such frameworks
aggregate. The Monarch Disease Ontology (MONDO)~\cite{mondo} and the
Human Phenotype Ontology (HPO)~\cite{hpo} provide established
vocabularies for disease and phenotype terms; we expect
curator-authored HPO terms to populate the phenotype-mapping fields of
our schema, as illustrated by the HPO-term suggestions in the Davis,
Nelson, and Gupta case studies (\cref{sec:workflow-flags}).

\paragraph{The coverage gap.}
None of the standards above was designed to treat as a first-class
concern the heterogeneous evidence patterns encountered in
basic-science and pre-clinical genetic publications: for example, a
Llgl1 conditional-knockout study that reports phenotypes at molecular,
cellular, and clinical scales in the same paper; a TTC21B study that
reports both case--control enrichment and an orthogonal functional
assay of 40 variants; or a metaGRS that spans 1.7 million variants
weighted into a single score. FHIR Evidence is shaped around
interventions; the GA4GH Statement model around single-variant
assertions; ECO around evidence codes for already-classified claims.
Each captures part of what such papers say, but none captures the
whole. The schema we introduce in the next section is an attempt to
represent that whole in a way that is compact enough to annotate
manually, machine-checkable enough to validate automatically, and
flexible enough to accommodate the heterogeneity the literature
actually produces.
