% =====================================================================
%  §3 — A Semantic Model of Genetic Evidence
%
%  Core classes, dimensional vocabulary, conditional activation
%  mechanism. Written to be self-contained; subsequent sections
%  reference back to the table and the formal snippets here.
% =====================================================================

\section{A Semantic Model of Genetic Evidence}
\label{sec:model}

The model comprises three parts: a small set of \emph{core classes}
(\cref{sec:model-classes}) that define what a unit of evidence is and
how evidence items compose; a \emph{dimensional vocabulary}
(\cref{sec:model-dimensions}) that describes each evidence item along
around twenty axes; and a \emph{conditional activation} mechanism
(\cref{sec:model-conditional}) that keeps the vocabulary compact by
requiring dimensions only when they are semantically appropriate.

\subsection{Core Classes}
\label{sec:model-classes}

Three abstract classes anchor the model:

\begin{description}
  \item[\ScientificEvidence] A unit of evidence, typically linked to a
        publication or dataset and described by a set of dimensions (see
        \cref{tab:dimensions-core,tab:dimensions-cond}). A single paper may
        give rise to multiple \ScientificEvidence{} items when it reports
        distinct evidential claims, for example a case--control association
        and an in vivo functional assay for the same variant.
  \item[\EvidenceVariable] A named quantity extracted from one or more
      \ScientificEvidence{} items, for example the gnomAD allele
      frequency of a variant, the LOD score in a family, or the odds
      ratio reported by a GWAS. Evidence variables are also annotated by
      dimensions, because the same quantity (for example, an odds ratio)
      may carry different epistemic weight depending on how it was
      obtained.
  \item[\EvidenceAssertion] A predicate over one or more
      \EvidenceVariable{} instances, intended to be machine-checkable
      and composable in decision workflows. Assertions may be positive
      or negative: they may state that a variant is associated with a
      phenotype, that it is \emph{not} associated, that its effect
      exceeds a threshold, and so on. The model does not restrict the
      syntactic form of the predicate; it only requires that every
      assertion be anchored to a specific source span in the underlying
      publication so that its provenance is auditable.
\end{description}

Specialising these three classes to the genetic domain yields
\GeneticEvidence, \GeneticEvidenceVariable, and
\GeneticEvidenceAssertion. The specialisations inherit the
provenance requirement and add the dimensional annotations
discussed next.

\subsection{Dimensions}
\label{sec:model-dimensions}

Each \GeneticEvidence{} instance is described using a dimensional
vocabulary that is summarised in \cref{tab:dimensions-core}
(always-required dimensions) and \cref{tab:dimensions-cond}
(representative conditional dimensions).\footnote{The full
enumeration of conditional dimensions, including those omitted
from \cref{tab:dimensions-cond} for space, is maintained in the
project repository at
\url{https://github.com/ForomePlatform/genetic-evidence-model/blob/master/schema/dimensions.md}.}
A small subset of dimensions is
\emph{always required}: Knowledge Domain, Method, Target Type,
Resolution, and Credibility. No well-formed genetic-evidence item is
fully specified without them. Together these dimensions tell a
downstream reader \emph{what kind} of evidence this is, \emph{how} it
was obtained, \emph{what} it is about, and \emph{how} credible it is
in context.

Two additional dimensions, Phenotype Scale and Variant Ascertainment,
were added to the core set during the annotation work described in
\cref{sec:workflow}. Phenotype Scale distinguishes evidence about
molecular phenotypes (for example, binding or activity) from cellular,
histological, organismal, or clinical phenotypes, and is independent
of the scale of the genetic target. Variant Ascertainment records how
the variant entered the study, such as being observed in affected
individuals or controls, drawn from a database of known polymorphisms,
or constructed synthetically. Both dimensions applied across all six
annotated publications and are revisited in \cref{sec:discussion}.

The remaining dimensions are \emph{conditional}: they are required, or
even meaningful, only when other dimensions take particular values. It
does not make sense to record mode of inheritance for a polygenic
score, or knockout type for a human case--control study. Instead of
marking such dimensions as ``optional'' and relying on convention, the
model attaches explicit activation conditions to each, formalised as
described in \cref{sec:model-conditional}. These conditions state the
situations in which a dimension must be present; outside those
situations it may be omitted. This makes the requirement status
transparent to both human curators and automated validators.
Credibility is an exception in form: unlike the other core dimensions
it is an open key--value field (\texttt{cohort\_size},
\texttt{replication\_cohort}, \texttt{multiple\_testing\_correction},
and so on) rather than a closed enumeration, because clinical and
research consumers weight credibility features differently and are
better served by preserving them than by collapsing them to a single
score.

\begin{table}[tbp]
\centering
\caption{Always-required dimensions of the \GeneticEvidence{} model.
These apply to every evidence item regardless of context. The final
two rows (Phenotype Scale and Variant Ascertainment) were added to
the always-required set during the annotation work in
\cref{sec:workflow}. KD abbreviates Knowledge Domain.}
\label{tab:dimensions-core}
\small
\begin{tabularx}{\linewidth}{@{}llllX@{}}
\toprule
Dimension              & Value type        & Card.    & Obj.       & Example values \\
\midrule
Knowledge Domain       & categorical       & multiple & obj.       & \dimval{Human Genetics}, \dimval{Population Genetics}, \dimval{Model Organism}, \dimval{Gene Function} \\
Method                 & categorical       & multiple & obj.       & \dimval{Statistical Genetics}, \dimval{In Vivo}, \dimval{In Vitro}, \dimval{Clinical Evidence} \\
Target Type            & categorical       & single   & obj.       & \dimval{Gene}, \dimval{Variant}, \dimval{Interval}, \dimval{Transcript} \\
Resolution             & categorical       & single   & obj.       & \dimval{Window}, \dimval{Gene}, \dimval{Functional Element}, \dimval{Position}, \dimval{Variant} \\
Credibility            & fuzzy key/value   & multiple & subj.      & \texttt{cohort\_size}, \texttt{replication}, \texttt{correction}, \ldots \\
Phenotype Scale        & categorical       & single   & obj.       & \dimval{Molecular}, \dimval{Cellular}, \dimval{Histological}, \dimval{Organismal}, \dimval{Clinical} \\
Variant Ascertainment  & categorical       & multiple & obj.       & \dimval{Observed in Cases}, \dimval{Observed in Controls}, \dimval{From Database}, \dimval{Synthetic} \\
\bottomrule
\end{tabularx}
\end{table}

\begin{table}[tbp]
\centering
\caption{Conditional dimensions of the \GeneticEvidence{} model.
Each is required only when its activation condition holds. KD
abbreviates Knowledge Domain; MT abbreviates Method; TT abbreviates
Target Type; $\supseteq$ denotes set membership for multi-valued
dimensions.}
\label{tab:dimensions-cond}
\small
\begin{tabularx}{\linewidth}{@{}lllX@{}}
\toprule
Dimension              & Value type   & Card.    & Activation condition \\
\midrule
Mode of Inheritance    & categorical  & single   & KD $\supseteq$ \dimval{Human Genetics} AND TT = \dimval{Gene} \\
Mendelian Segregation  & boolean      & single   & KD $\supseteq$ \dimval{Human Genetics} AND TT = \dimval{Gene} \\
Penetrance             & categorical  & single   & KD $\supseteq$ \dimval{Human Genetics} AND TT = \dimval{Variant} \\
Measurement Target     & categorical  & multiple & KD $\supseteq$ \dimval{Gene Function} \\
Gene Relation          & categorical  & single   & KD $\supseteq$ \dimval{Gene Function} \\
Organism               & categorical  & multiple & MT $\supseteq$ \dimval{In Vivo} OR KD $\supseteq$ \dimval{Model Organism} \\
Knockout Type          & categorical  & single   & KD $\supseteq$ \dimval{Model Organism} \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Conditional Activation}
\label{sec:model-conditional}

The conditional activation mechanism is the least conventional part of
the model and is what keeps the vocabulary compact without sacrificing
expressive power. Instead of a fixed schema in which every
\GeneticEvidence{} item has every dimension (with many dimensions
marked as optional and left empty for most items), the model states
that certain dimensions are required \emph{exactly when} specified
conditions over other dimensions hold. This mirrors how domain experts
reason: a human curator reading a paper about an X-linked recessive
pedigree knows to ask about mode of inheritance because the evidence
makes that question meaningful, not because a global schema demands it.

We formalise activation conditions as SHACL shapes using SPARQL-based
constraints, which are the idiomatic SHACL mechanism for expressing
requirements that cannot be captured by simple cardinality annotations
on a single property. Two representative shapes illustrate the
pattern; we show only the \texttt{sh:sparql} body of each and elide
the outer NodeShape wrapper for space. The full shapes, together
with the base \texttt{GeneticEvidenceShape} and the simpler
\texttt{variant\_ascertainment} conditional rule, are distributed
with annotated walkthrough in the project repository.\footnote{\url{https://github.com/ForomePlatform/genetic-evidence-model/blob/master/schema/examples.md}}

The first example encodes a conjunction over two axes of
classification: \emph{mode of inheritance} is required when the
Knowledge Domain contains \dimval{Human Genetics} \textbf{and} the
Target Type is \dimval{Gene} (\cref{lst:shacl-moi}).

\begin{lstlisting}[language=yaml, caption={Compact form of the conditional-activation shape for \emph{mode of inheritance}: the outer NodeShape wrapper is elided. The two triple patterns in \texttt{WHERE} form the conjunction; \texttt{FILTER NOT EXISTS} identifies instances that satisfy the condition but lack the required dimension.}, label={lst:shacl-moi}]
sh:sparql [ sh:select """
  SELECT $this WHERE {
    $this gem:knowledgeDomain gem:HUMAN_GENETICS .
    $this gem:targetType      gem:GENE .
    FILTER NOT EXISTS { $this gem:modeOfInheritance ?m }
  }""" ] .
\end{lstlisting}

The second example encodes a cross-axis disjunction:
\emph{organism} is required when the Method contains
\dimval{In Vivo Experiment} \textbf{or} the Knowledge Domain contains
\dimval{Model Organism} (\cref{lst:shacl-org}). The two disjuncts
refer to different axes, one a Method value and the other a
Knowledge Domain value, illustrating that activation conditions can
span the dimensional vocabulary rather than being confined to a
single axis.

\begin{lstlisting}[language=yaml, caption={Compact form of the conditional-activation shape for \emph{organism}: the disjunction is expressed by a SPARQL \texttt{UNION} whose two branches refer to different dimensions.}, label={lst:shacl-org}]
sh:sparql [ sh:select """
  SELECT $this WHERE {
    { $this gem:method          gem:IN_VIVO_EXPERIMENT . }
    UNION
    { $this gem:knowledgeDomain gem:MODEL_ORGANISM . }
    FILTER NOT EXISTS { $this gem:organism ?o }
  }""" ] .
\end{lstlisting}

In each case, the \texttt{SELECT} returns every \GeneticEvidence{}
instance that satisfies the activation condition but lacks the
required dimension; the validator reports those instances as
constraint violations with a localised diagnostic. The small
additional cost of writing each conditional rule explicitly is
repaid in downstream clarity: a reader of an annotation file can
see \emph{why} a particular dimension is required or absent, and a
validator can check conformance automatically.

\Cref{sec:workflow} returns to conditional activation in the
context of the annotation workflow; there we report that all six
publications we annotated could be placed into the model with
conditional activation behaving as designed.
