% =====================================================================
%  §4 — Worked Examples from the Literature
%
%  Six-paper corpus table, two featured examples (Duerr 2006, Inouye
%  2018), summary pointer to GitHub for the full annotations of all
%  six.
% =====================================================================

\section{Worked Examples from the Literature}
\label{sec:examples}

To demonstrate the model's coverage and to expose the places where it
strains, we applied it to six published genetic-evidence papers
spanning the major epistemic shapes encountered in literature-based
variant interpretation (\cref{tab:corpus}). Four of the six were
annotated manually from the curator's PDF highlights and callouts and
are treated as ground truth. The remaining two were drafted by an AI
assistant working from the PDF text alone, with the curator's review
serving as the evaluation signal. All six annotations are available in
the project repository; this section features two in detail.

\begin{table}[tbp]
\centering
\caption{The six annotated publications and the epistemic shape each
exercises. Manual annotations are treated as ground truth; AI-drafted
annotations are evaluated against curator review. GE = \GeneticEvidence{}
items produced per paper after decomposition at natural-claim granularity.}
\label{tab:corpus}
\small
\begin{tabularx}{\linewidth}{@{}lllllX@{}}
\toprule
Paper & Source & GE items & Credibility & Role & Notes \\
\midrule
Jossin 2017 (Llgl1)   & manual       & 6 & Very high & molecular mechanism           & multi-scale phenotype; motivates \emph{phenotype\_scale} \\
Davis 2011 (TTC21B)   & manual       & 6 & High      & breadth exemplar              & motivates \emph{variant\_ascertainment} \\
Nelson 1992 (CD18)    & manual       & 3 & Medium    & classical molecular genetics  & single-gene functional; no extensions surfaced \\
Gupta 2015 (ATP6AP2)  & manual       & 2 & Low       & low-credibility edge case     & schema fits cleanly \\
Duerr 2006 (IL23R)    & AI-drafted   & 6 & High      & clean GWAS exemplar           & featured below \\
Inouye 2018 (metaGRS) & AI-drafted   & 5 & High/Med  & model-extension stress test   & featured below \\
\bottomrule
\end{tabularx}
\end{table}

We feature two of the papers here. Duerr et al.\ 2006
(\cref{sec:ex-duerr}) is a textbook GWAS whose evidence shape the
model accommodates cleanly; it shows the machinery in ordinary
operation. Inouye et al.\ 2018 (\cref{sec:ex-inouye}) is a polygenic
risk score study whose evidence shape deliberately strains the model;
its annotation surfaces a cluster of candidate extensions that we
return to in \cref{sec:discussion}. The other four annotations
(Jossin, Davis, Nelson, Gupta) are described in sufficient detail in
the repository's \texttt{annotations/} directory, and we mention them
below only to illustrate specific methodological points.

\subsection{A Genome-Wide Association Study: Duerr et al.\ 2006}
\label{sec:ex-duerr}

Duerr and colleagues identified the \textit{IL23R} association with
inflammatory bowel disease by typing 308{,}332 autosomal SNPs in 547
non-Jewish European-ancestry cases with ileal Crohn's disease against
548 matched controls, following genome-wide significance with
independent replication in a Jewish-ancestry cohort and family-based
transmission disequilibrium testing in 883 nuclear families. The
paper's central finding---that the Arg381Gln coding variant
rs11209026 confers protection against Crohn's disease with an odds
ratio of 0.26 in the discovery cohort---is the kind of result the
model was designed to represent: a variant-level claim grounded in
population-genetic methodology, with explicit cohort descriptors, a
multiple-testing adjustment, and external replication.

Our annotation decomposes the paper into six \GeneticEvidence{} items
covering, respectively, the GWAS discovery, the Jewish-cohort
replication, the family-based transmission analysis, a conditional
analysis showing that the IL23R signal comprises more than one
independent variant, a negative result at three related IL-23 pathway
genes, and cited mouse-model evidence supporting a therapeutic
rationale. All six items fit the schema without invoking any
conditional dimension that was not already part of the model, and
only one candidate extension (a direction-of-effect dimension for
protective versus risk-conferring alleles) was surfaced, which awaits
a second-paper confirmation before promotion.

Listing~\ref{lst:duerr} shows the annotation of the paper's primary
assertion---the discovery of rs11209026's association---abridged to
fit the page. The full annotation, including all six evidence items
and their source-anchored sub-assertions, is available in the
project repository.

\begin{lstlisting}[language=yaml, caption={Abridged \GeneticEvidence{} annotation for the primary discovery in Duerr et al.\ 2006. Source-span and credibility fields are retained; some sub-fields (e.g., \texttt{special\_considerations}) are elided for space. \texttt{source: ai\_drafted} marks the item as AI-generated and pending curator review.}, label={lst:duerr}]
- id: GE-1
  label: "GWAS discovery of rs11209026 in IL23R"
  source: ai_drafted
  knowledge_domain: [POPULATION_GENETICS]
  method: [STATISTICAL_GENETICS]
  target_type: VARIANT
  target: "rs11209026 (p.Arg381Gln) in IL23R"
  resolution: VARIANT
  variant_ascertainment: [OBSERVED_IN_CASES, OBSERVED_IN_CONTROLS]
  phenotype_scale: CLINICAL
  subdomain: "GWAS"
  credibility: HIGH
  credibility_comment: >
    Genome-wide significance after Bonferroni correction
    (corrected P = 1.56e-3). OR = 0.26 (95% CI 0.15-0.43).
    Replication follows in GE-2.
  assertions:
    - id: GE-1.A1
      statement: "rs11209026 is genome-wide significantly associated with ileal Crohn's disease"
      source_span:
        page: 1
        key_phrase: "rs11209026 (P = 5.05e-9, corrected P = 1.56e-3)"
    - id: GE-1.A2
      statement: "The Gln allele is protective (OR = 0.26, 95% CI 0.15-0.43)"
      source_span:
        page: 2
        key_phrase: "glutamine allele appears to protect against development of CD"
\end{lstlisting}

Two things in the listing are worth calling attention to. First, the
\texttt{variant\_ascertainment} field records that rs11209026 was
observed in both cases and controls, which it was---the
case--control design ascertains variants in both groups by
construction. This instance of the dimension is what justifies its
promotion to the core schema: the Duerr annotation exercises it, as
does the Davis annotation (patient-observed vs.\ functionally-tested
variants), as does every variant-level annotation in the corpus.
Second, each assertion carries a source span pointing to a specific
page and a short identifying phrase from the paper, which is what
makes the annotation auditable---a reader disagreeing with our
interpretation can check directly.

\subsection{A Polygenic Risk Score: Inouye et al.\ 2018}
\label{sec:ex-inouye}

The polygenic risk score literature stresses the model in a different
direction. Inouye and colleagues constructed a meta-genomic risk
score (metaGRS) for coronary artery disease from 1{,}745{,}180 SNPs,
weighted by effect size, combining three prior polygenic scores; they
validated the score externally in 482{,}629 UK Biobank participants,
reported a hazard ratio per standard deviation of 1.71, demonstrated
substantial independence from conventional risk factors, and showed
that the score's predictive power persists in medicated
subpopulations. Every one of these claims is a genuine piece of
genetic evidence, and every one of them stretches the model: the
target is not a gene or a variant but a derived score, the
resolution is not a locus but the whole genome in aggregate, and the
measurement targets are epidemiological (hazard ratio, C-index,
cumulative incidence) rather than molecular. The annotation surfaces
six candidate extensions as a direct consequence, all of which we
return to in \cref{sec:discussion}.

Listing~\ref{lst:inouye} shows the UK Biobank validation evidence
item. Fields whose values required the use of placeholder
enumeration entries are accompanied by
\texttt{candidate\_extension\_exercised} tags, which link to the
proposed extensions recorded in the annotation's
\texttt{candidate\_extensions} block.

\begin{lstlisting}[language=yaml, caption={Abridged \GeneticEvidence{} annotation for the UK Biobank external validation of the metaGRS. The candidate-extension tags identify the fields whose current values are placeholders from the existing enumerations, with the proposed new enumeration values recorded in the full annotation. This makes the annotation's strain on the schema explicit and auditable.}, label={lst:inouye}]
- id: GE-2
  label: "External validation of metaGRS in UK Biobank"
  source: ai_drafted
  knowledge_domain: [POPULATION_GENETICS]
  method: [STATISTICAL_GENETICS]
  target_type: GENE       # placeholder; candidate: SCORE
  candidate_extension_exercised: [CE-IN1, CE-IN2, CE-IN3, CE-IN4]
  target: "metaGRS"
  resolution: GENE        # placeholder; candidate: WHOLE_GENOME_AGGREGATE
  variant_ascertainment: [FROM_DATABASE]
  phenotype_scale: CLINICAL
  credibility: HIGH
  credibility_comment: >
    Prospective cohort of 482,629 individuals independent of training
    set. <5% non-European ancestry limits generalisability.
  assertions:
    - id: GE-2.A1
      statement: "Hazard ratio per standard deviation of metaGRS is 1.71"
      measurement_type: HAZARD_RATIO   # candidate: CE-IN4
      source_span:
        page: 1
        key_phrase: "HR for CAD was 1.71 (95% CI 1.68 to 1.73) per SD"
    - id: GE-2.A3
      statement: "metaGRS C-index exceeds every conventional risk factor"
      measurement_type: C_INDEX        # candidate: CE-IN4
      source_span:
        page: 1
        key_phrase: "C-index = 0.623 (95% CI 0.615-0.631) for incident CAD"
\end{lstlisting}

The four candidate extensions flagged inline---\texttt{CE-IN1}
through \texttt{CE-IN4}---propose, respectively, a \dimval{Score}
value for Target Type, a \dimval{Whole Genome Aggregate} value for
Resolution, a new \emph{target composition} dimension distinguishing
single versus aggregate targets, and an extension of the measurement
target enumeration to cover epidemiological quantities such as
\dimval{Hazard Ratio} and \dimval{C-Index}. Two further extensions
(CE-IN5 on structured cohort descriptors, CE-IN6 on distinguishing
natural from derived-artifact targets) are surfaced elsewhere in the
annotation. The important methodological point is that the model's
strain is \emph{legible}: rather than silently forcing a polygenic
score into a variant-level field and producing a misleading
annotation, the schema requires the curator (human or AI) to mark
the forced fit and propose an extension. \cref{sec:workflow}
describes the promotion rule that governs when such candidates become
first-class, and \cref{sec:discussion} reports what we learned about
the model from the pattern of extensions that the six papers
surfaced as a group.
