% =====================================================================
%  §5 — Human-AI Annotation Workflow and Evaluation
%
%  The ICBO-theme hook section. Describes the two-track workflow
%  (manual ground truth vs AI-drafted), coverage and agreement
%  metrics computed from the six YAMLs, the flag taxonomy with
%  worked examples, and the extension-promotion rule with the
%  CE-DU2 retraction as a positive example.
%
%  Honest about the small corpus: called a "feasibility study" in
%  §5.4 not a benchmark.
% =====================================================================

\section{Human--AI Annotation Workflow and Evaluation}
\label{sec:workflow}

The model is a schema, but a schema is only as useful as the
annotations it supports. We designed and piloted an annotation
workflow whose central commitment is that every annotation be
\emph{auditable}: an assertion anchored to the specific text that
justifies it, a dimension value justifiable against the schema's
activation conditions, and a reviewer flag on any judgment call that
a downstream consumer might contest. The same commitment extends to
AI assistance: when an AI system drafts an annotation, its
uncertainties and assumptions are not silently absorbed but recorded
alongside the annotation for curator review.

This section describes the protocol (\cref{sec:workflow-protocol}),
reports coverage and flag statistics from the six-paper pilot
(\cref{sec:workflow-coverage,sec:workflow-flags}), and examines the
extension-promotion rule that determines when a candidate schema
extension becomes first-class (\cref{sec:workflow-promotion}). We
treat the pilot as a \emph{feasibility study} rather than a
benchmark: six papers and one annotator are not enough to make
claims about inter-annotator agreement or generalisability. What the
pilot does show is that the workflow's epistemic commitments are
practical to meet and that the schema's strain points are surfaced
rather than hidden.

\subsection{Protocol}
\label{sec:workflow-protocol}

Annotations are authored in two tracks.

\paragraph{Manual track.}
For four of the six publications (Jossin, Davis, Nelson, Gupta) a
human curator read the paper and marked the relevant passages
directly in the PDF as highlights and sticky-note callouts. A short
extraction script (provided in the project repository) pulls the
highlights and callouts into a structured JSON record. The curator's
judgements---expressed as a summary table of dimension values and
as free-text callouts attached to specific passages---are then
transformed into a YAML annotation that conforms to the schema. Where
the transformation requires a judgment that the curator did not
make explicit (for example, normalising the free-text value
``localisation and protein expression'' to the enumeration values
\dimval{Localization} and \dimval{Existence}), the transformation is
recorded inline as a \texttt{normalization\_note} with a
\texttt{type} tag (\texttt{ai\_normalization}, in this case) so that
the curator can audit the transformation later.

\paragraph{AI-drafted track.}
For two publications (Duerr, Inouye) the annotation was drafted
directly by an AI assistant\footnote{Claude Opus 4.7 Adaptive.}
working from the PDF text. The AI was given the schema (as
\texttt{dimensions.md}), the ground-truth annotations of the four
manual papers for reference, and the PDFs themselves. It produced
YAML annotations in the same format. Where the AI judged that a
mapping was uncertain or that a judgement call ought to be verified
by a human reviewer, it recorded the uncertainty inline (see
\cref{sec:workflow-flags}). The curator then reviewed each AI-drafted
annotation assertion-by-assertion, flagging any that misrepresented
the paper or misused the schema.

\paragraph{Source anchoring.}
In both tracks, every
\GeneticEvidenceAssertion{} carries a \texttt{source\_span} pointing
to a specific page of the publication and a short key phrase
(typically under fifteen words) from the underlying text. The phrase
is chosen to be long enough that a reader can locate the passage by
search and short enough to stay within fair-use norms for
copyrighted publisher material. Source anchoring is not optional:
an annotation without source spans is rejected at validation time.
Across the six-paper pilot, 65 of 65 assertions (100\%) are
source-anchored; this is by construction, not by observation.

\subsection{Coverage}
\label{sec:workflow-coverage}

\Cref{tab:coverage} reports, for each schema dimension, the
fraction of applicable \GeneticEvidence{} items in which the
dimension is populated with a real value. The denominator is the
number of items where the dimension's activation condition holds;
for the always-required dimensions the denominator is always 28
(the total number of \GeneticEvidence{} items produced across the
corpus).

\begin{table}[tbp]
\centering
\caption{Dimension population rates across 28 \GeneticEvidence{}
items from six papers. The denominator for each conditional
dimension is the number of items whose activation condition holds.
A rate of 100\% indicates that every applicable item has the
dimension populated.}
\label{tab:coverage}
\small
\begin{tabular}{@{}llcc@{}}
\toprule
Dimension              & Class        & Populated & Applicable \\
\midrule
Knowledge Domain       & always       & 28        & 28 \\
Method                 & always       & 28        & 28 \\
Target Type            & always       & 28        & 28 \\
Resolution             & always       & 28        & 28 \\
Credibility            & always       & 28        & 28 \\
Phenotype Scale        & always       & 28        & 28 \\
Variant Ascertainment  & cond.\ on TT & 21        & 21 \\
\midrule
Mode of Inheritance    & cond.        & 0         & 1 \\
Mendelian Segregation  & cond.        & 0         & 1 \\
Penetrance             & cond.        & 2         & 4 \\
Measurement Target     & cond.        & 10        & 11 \\
Gene Relation          & cond.        & 1         & 11 \\
Organism               & cond.        & 7         & 8 \\
Knockout Type          & cond.        & 1         & 6 \\
\bottomrule
\end{tabular}
\end{table}

The always-required dimensions are fully populated, which is by
construction: the schema rejects annotations that leave them empty.
The conditional-dimension population rates are more informative.
Three patterns are worth remarking on.

First, \emph{high-population} dimensions (Measurement Target at
10/11, Organism at 7/8) indicate that the corresponding activation
condition is well-calibrated: when the rule says the dimension
should apply, it almost always does. Second, \emph{zero-population}
dimensions (Mode of Inheritance at 0/1, Mendelian Segregation at
0/1) indicate boundary cases. The single applicable item for each
is Duerr's family-based transmission disequilibrium analysis, which
is a complex-trait genetic study; the annotator recorded explicit
uncertainty about whether the enumeration value \dimval{Complex}
ought to be a first-class member of the Mode-of-Inheritance
enumeration (\cref{sec:workflow-flags}, \textit{ai\_uncertainty} on
Duerr GE-3). Third, \emph{low-but-nonzero} dimensions (Gene
Relation at 1/11) reveal schema shortcomings: in the eleven
\dimval{Gene Function} evidence items, only one had a gene relation
that fit the current three-value enumeration
(\dimval{X regulates Y}, \dimval{X inhibits Y},
\dimval{X has same function as Y}); the remaining ten mostly
involved physical binding or trafficking, which the enumeration
does not cover. This observation motivates the candidate extension
CE-J3 surfaced in the Jossin annotation
(\cref{sec:discussion}).

These coverage statistics are internally consistent with the
activation rules: no dimension is populated where the rule says it
should not be, and the few places where a rule-required dimension
is left empty are each accompanied by an explicit flag. Coverage
alone does not tell us whether the annotations are \emph{correct}
(that would require a second annotator), only whether the schema's
structural requirements are met.

\subsection{Flag Taxonomy}
\label{sec:workflow-flags}

The two tracks use different flag vocabularies because they
represent different epistemic situations.

\paragraph{Manual track: three flag types.}
When a curator-authored annotation diverges from the curator's own
summary table, or when an AI reviewer pass over the manual
annotation detected a mapping issue, we record one of three flag
types. A \texttt{reviewer\_disagreement} marks a clear factual
correction: in the Davis 2011 annotation, the curator's summary
listed the organism set as \{mouse, zebrafish\}, but the paper's
rat-retinal electroporation experiment was passed over in the
summary, and the AI reviewer proposed adding \dimval{Rat} to the
set. A \texttt{reviewer\_suggestion} marks an optional addition
whose inclusion depends on curator preference: HPO terms for
patient phenotypes are a recurring case, flagged in Davis, Nelson,
and Gupta. A \texttt{reviewer\_query} marks a case where the
mapping itself is unclear and curator input is needed: the Nelson
2011 summary table gave
\texttt{specificity\_of\_phenotype = SNV}, which is semantically
odd (SNV is a variant type, not a phenotype-specificity
descriptor), and the annotation preserves the original value with
a query rather than silently correcting it.

\paragraph{AI-drafted track: two flag types.}
When no prior ground truth exists, an AI-drafted annotation has
nothing to \emph{disagree} with. The AI track therefore uses a
narrower vocabulary: an \texttt{ai\_uncertainty} marks a case where
the AI was unsure about a schema mapping and would prefer human
input (Duerr's \dimval{Complex} value for Mode of Inheritance,
above, is one); an \texttt{ai\_assumption} marks a judgement call
the AI made without strong evidence, which the reviewer should
verify. An example of the latter is Duerr's GE-6, which collects
citations to prior mouse-model work that the paper's Discussion
leverages to motivate the therapeutic claim; the AI drafted this
as a \GeneticEvidence{} item but flagged that it is not clear
whether evidence \emph{cited} by a paper belongs in the annotation
of \emph{that} paper, or whether citation chains require a
separate representational layer. This is a judgment call we think
worth recording but not resolving silently.

\Cref{tab:flags} summarises the flag counts across the six
annotations.

\begin{table}[tbp]
\centering
\caption{Reviewer and AI flags across the six-paper pilot. Each
cell is a count per paper. Manual-track flags (query, suggestion,
disagreement) apply to the four manually curated papers; AI-track
flags (uncertainty, assumption) apply to the two AI-drafted papers.
Em dashes indicate inapplicable flag--paper combinations.}
\label{tab:flags}
\small
\begin{tabular}{@{}lccccc@{}}
\toprule
Paper      & Query & Suggestion & Disagree & AI unc. & AI assum. \\
\midrule
Jossin     & 3     & 1          & 0        & ---      & ---        \\
Davis      & 2     & 1          & 1        & ---      & ---        \\
Nelson     & 0     & 3          & 0        & ---      & ---        \\
Gupta      & 0     & 1          & 0        & ---      & ---        \\
Duerr      & ---   & ---        & ---      & 3        & 2          \\
Inouye     & ---   & ---        & ---      & 0        & 2          \\
\midrule
Total      & 5     & 6          & 1        & 3        & 4          \\
\bottomrule
\end{tabular}
\end{table}

The totals themselves are less informative than the distribution.
Disagreements are rare (one across the four manual papers), which
is expected given that the annotator also authored the summary
being compared against. Suggestions and queries are the workhorses:
six of each, typically for edge cases (HPO-term candidates,
knockout-type vs.\ mode-of-inheritance scoping, ``SNV in the
phenotype slot'') where human judgement beyond the schema is
needed. For the AI-drafted papers, Duerr generated three
uncertainties (knowledge-domain scoping for GE-1, mode-of-inheritance
enumeration for GE-3, the cited-vs-generated boundary for GE-6) and
two assumptions (source-phrase selection and the placeholder choice
for GE-3); Inouye generated no uncertainties at all (the AI
identified each stress-test point as a candidate-extension
rather than an uncertain mapping) and two assumptions about
measurement-target encoding.

\subsection{The Extension-Promotion Rule}
\label{sec:workflow-promotion}

The workflow's most substantive methodological commitment is its
rule for when a candidate schema extension becomes a first-class
part of the model. The rule is simple and deliberately
conservative: a gap surfaced in exactly one paper is recorded in
that paper's \texttt{candidate\_extensions} block but does not
change the schema; a gap surfaced \emph{independently} in a second
paper is promoted to a first-class dimension or enumeration
value. The rationale is that the biomedical literature is
heterogeneous enough that any given paper can surface
idiosyncrasies that do not generalise, and the schema should only
grow in response to convergent evidence.

In the six-paper pilot the rule has fired twice. \emph{Phenotype
Scale} was first proposed by the Jossin annotation (to
distinguish the paper's three ``phenotype'' callouts at molecular,
histological, and organismal scales) and then independently
exercised by Davis, Nelson, and all subsequent papers; it was
promoted to the always-required set and is now part of
\cref{tab:dimensions-core}. \emph{Variant Ascertainment} was first
proposed by the Davis annotation (to distinguish variants observed
in patients from variants tested because they appeared in a
database) and then independently exercised by Nelson, Duerr,
and Inouye; it was promoted conditionally on
\mbox{\dimval{Target Type}~=~\dimval{Variant}} and is also now
in \cref{tab:dimensions-core}. Thirteen candidate extensions
remain unpromoted, pending either a second encounter or an
explicit curator decision.

A complementary positive finding from the pilot is a candidate
extension that the rule \emph{correctly} blocked. The Duerr
annotation initially proposed a new dimension for handling
negative-polarity assertions (GE-5 asserts ``no significant
association at IL12RB1, IL23A, or IL12B''), on the reasoning
that the schema seemed oriented toward positive claims. On
review the curator noted that an \EvidenceAssertion{} is a
predicate, and a predicate can state the absence of something
just as readily as its presence; no new dimension was needed.
The candidate was retracted before any schema change occurred.
This is the kind of false-positive extension that a
conservative promotion rule is designed to catch: had the rule
permitted single-paper promotion, the schema would now carry
a redundant polarity field that would never have been exercised
again. The retraction is preserved in the project repository
for methodological completeness.

\subsection{Discussion of Methodology}
\label{sec:workflow-discussion}

Three features of the workflow are worth making explicit before
we turn to broader implications in \cref{sec:discussion}.

First, the workflow treats schema gaps as \emph{data} rather than
as problems to be hidden. When an AI-drafted annotation cannot fit
a claim into the existing enumeration, it records the strain
inline with a candidate-extension tag and continues. When a
curator-authored annotation finds the summary-table value at odds
with the schema, the disagreement is preserved alongside the
resolution. The alternative---silently forcing the claim into the
nearest available enumeration value, or quietly extending the
enumeration without marking the change---produces annotations
that are cleaner but lose information about where the schema
strains.

Second, the workflow's two-track structure separates two
different epistemic roles that the literature tends to conflate.
A manually-curated annotation is a piece of primary knowledge
work: it reflects a human reading of the source. An AI-drafted
annotation is a piece of \emph{secondary} knowledge work: it
reflects a machine reading that is valid only insofar as a human
reviewer assents. The flag vocabularies differ because the
epistemic situations differ; conflating them---treating
AI-drafted annotations as if they were ground truth, or
treating manually-curated annotations as if they required
assumption tags---loses information about the provenance of each
claim.

Third, the six-paper pilot is deliberately small. We did not set
out to benchmark the workflow against inter-annotator agreement
metrics computed over dozens of papers; we set out to verify that
the schema's structural requirements are practical to meet, and
that the workflow's flag vocabularies capture the kinds of
disagreement that arise in practice. We report below on the
specific model-extension findings that the pilot surfaced
(\cref{sec:discussion}), and flag as future work the question of
whether the workflow scales to larger corpora and multiple
annotators.
