% =====================================================================
%  §6 — Discussion
%
%  Structure:
%    §6 opening:     what the paper did and for whom
%    §6.1:           ICBO 2026 theme alignment
%    §6.2:           Model extensions surfaced by annotation
%                    (the methodology payoff — grouped candidate table)
%    §6.3:           Limitations
%    §6.4:           Future work
%
%  Avoids restating the model (§3's job) and over-claiming
%  generalisability from a 6-paper pilot.
% =====================================================================

\section{Discussion}
\label{sec:discussion}

We introduced a semantic model for genetic evidence
(\cref{sec:model}), applied it to a heterogeneous corpus of six
publications spanning classical molecular genetics, population
genetics, model-organism biology, and polygenic risk scoring
(\cref{sec:examples}), and piloted a two-track annotation workflow
with explicit flag vocabularies for both human-curated and
AI-drafted annotations (\cref{sec:workflow}). Two features of the
model are worth recapitulating as the framing for the discussion
that follows. First, the model's always-required vocabulary is
small (seven dimensions) and its conditional-activation mechanism
keeps the extended vocabulary compact without sacrificing
expressive power: an item describing a \dimval{Model Organism}
knockout is required to declare its organism and knockout type,
while an item describing a human genetic association is not.
Second, every assertion in every annotation is anchored to a
specific source span in its underlying publication, making each
annotation independently auditable. These two features, we argue,
together distinguish the proposal from the larger class of evidence
schemas that either fix a flat vocabulary or rely on convention
rather than formal constraint for provenance.

\subsection{Alignment with the ICBO 2026 Theme}
\label{sec:disc-icbo}

The conference call asks for contributions that advance
trustworthy, interpretable, and AI-ready semantic infrastructure.
The present work contributes to each of these in a concrete way
that was not invented to match the theme but reflects decisions
made during the corpus pilot.

\emph{Trustworthiness.} Every schema conformance requirement is
expressed formally as a SHACL shape, and every activation
condition is machine-checkable. An annotation that violates a
conditional-activation rule is flagged automatically; the rule
itself is auditable from the schema file rather than from a style
guide. The source-anchoring requirement makes each assertion
independently verifiable against its source; a reader disagreeing
with any assertion can locate the specific passage the annotation
rests on without tracing through the paper.

\emph{Interpretability.} The flag vocabularies of
\cref{sec:workflow-flags} make the provenance of each judgment
explicit: a reader encountering a value in an annotation can tell
whether it was drawn from the author's summary, proposed by a
reviewer, added as an AI assumption, or left deliberately
unpopulated. No value in a conforming annotation is orphaned; each
either fits a schema constraint or carries a tag explaining why it
does not.

\emph{AI-readiness.} The annotations are machine-actionable not
because they are simple but because they carry their own
metadata. An automated consumer that needs only high-credibility
human-genetics evidence can filter on the relevant dimensions;
one that needs to reason about cross-level variant descriptions
(HGVS cDNA vs.\ genomic coordinates) can route to
\texttt{special\_considerations} entries keyed to that concern.
The schema is readable by both humans and validators, and the
structure is stable across papers that span an order of magnitude
in sample size and half a century in publication date.

A broader way to frame the same point is that the schema
is designed to make literature review tractable as a
\emph{semantic parsing} problem: the mapping of natural-language
claims in a publication to a shared, source-anchored
representation. Modern large language models have become
substantially more capable at semantic parsing than they were
only a few years ago, and the flag vocabularies of
\cref{sec:workflow-flags} were in fact designed with a parser's
three possible outcomes in mind---a successful mapping,
a request for human disambiguation, or a proposal to extend the
target language. That said, semantic parsing of the biomedical
literature is not something a language model can perform in
isolation, and we do not claim otherwise: the target language
into which each claim is parsed is an ontology, and the quality
of the parse depends on the quality of the ontological grounding.
The proposed schema is one such target, partial and evolving,
whose development rests on decades of prior work in biomedical
ontology (ECO, SEPIO, HPO, IAO, OBI, VRS, and others).
We see the present contribution as one increment along a
longer path, not as a replacement for any of the work it
builds on.

\subsection{Model Extensions Surfaced by Annotation}
\label{sec:disc-extensions}

The most substantive methodological finding of the pilot is that
applying the schema to a small but heterogeneous corpus surfaced
specific, grounded proposals for how the schema should evolve.
Two dimensions---\emph{Phenotype Scale} and \emph{Variant
Ascertainment}---were proposed by individual papers, independently
exercised by others, and consequently promoted to first-class
status in the always-required vocabulary
(\cref{tab:dimensions-core}). Thirteen additional candidate
extensions were surfaced by individual annotations and recorded
in their \texttt{candidate\_extensions} blocks; under the
promotion rule of \cref{sec:workflow-promotion} none has yet been
promoted, because no single candidate has been independently
exercised by a second paper in the six-paper pilot. These
candidates are listed in \cref{tab:candidates}, grouped by what
they suggest about the model's initial scope.

\begin{table}[tbp]
\centering
\caption{Candidate schema extensions surfaced by the corpus
pilot, grouped by concern. None has yet been promoted to
first-class under the two-papers rule. Entries are keyed by the
paper that surfaced them (J~=~Jossin, D~=~Davis, N~=~Nelson,
DU~=~Duerr, IN~=~Inouye; Gupta surfaced none).}
\label{tab:candidates}
\small
\begin{tabularx}{\linewidth}{@{}llX@{}}
\toprule
ID      & Source & Proposed extension \\
\midrule
\multicolumn{3}{@{}l}{\emph{Finer structural resolution}} \\
CE-J1   & J      & \dimval{Protein Subdomain} value for Resolution \\
CE-J2   & J      & \dimval{Interaction} and \dimval{Complex} values for Target Type \\
CE-J3   & J      & \dimval{X physically binds Y} and \dimval{X regulates trafficking of Y} values for Gene Relation \\
CE-N1   & N      & Cross-level variant description (gDNA vs.\ cDNA vs.\ protein); resolvable via reference to HGVS / GA4GH VRS \\
\midrule
\multicolumn{3}{@{}l}{\emph{Polygenic-score machinery}} \\
CE-IN1  & IN     & \dimval{Score} value for Target Type \\
CE-IN2  & IN     & \dimval{Whole Genome Aggregate} value for Resolution \\
CE-IN3  & IN     & \emph{target\_composition} dimension: \dimval{Single} / \dimval{Aggregate} / \dimval{Composite} \\
CE-IN4  & IN     & Epidemiological Measurement Target values: \dimval{Hazard Ratio}, \dimval{Odds Ratio}, \dimval{C-Index}, \dimval{AUROC}, \ldots \\
CE-IN5  & IN     & Structured \emph{cohort} sub-block (n\_total, ancestry, follow-up, ascertainment) \\
CE-IN6  & IN     & \emph{derived\_artifact} flag distinguishing natural from constructed targets \\
\midrule
\multicolumn{3}{@{}l}{\emph{Effect-size direction and curation}} \\
CE-DU1  & DU     & \emph{direction\_of\_effect} dimension: \dimval{Risk}/\dimval{Protective}/\dimval{Neutral} \\
CE-D1   & D      & Ordinal priority on multi-valued \emph{knowledge\_domain} \\
CE-D2   & D      & \emph{curator\_critique} sub-block on \EvidenceAssertion{} \\
\bottomrule
\end{tabularx}
\end{table}

The grouping itself is the interpretation. The first cluster
(finer structural resolution) is what the model would need to
accommodate papers that reason below the gene level---to
protein subdomains, to interactions, to cross-level variant
descriptions. This is largely solved by external vocabularies we
would simply reference rather than reinvent: PROTEIN\_SUBDOMAIN
could be grounded in InterPro; the HGVS / GA4GH VRS ecosystem
already provides the cross-level variant identifiers that CE-N1
identifies as missing. The second cluster (polygenic-score
machinery) is different in character: it is concentrated in a
single paper and reflects a genuine category shift from
variant-level to aggregate-score-level evidence. Whether these
candidates become first-class dimensions or whether polygenic
scores warrant a distinct \GeneticEvidence{} subtype is a
question that would be settled by annotating additional
polygenic-score papers, which the two-papers promotion rule
anticipates. The third cluster (effect-size direction and
curation) is the smallest and most heterogeneous: each candidate
responds to a specific curator judgement that the current schema
does not have a place for.

A negative finding is worth naming as well. The Gupta annotation
surfaced no candidate extensions. This was not because the paper
is trivial---it reports a family with an X-linked splice-site
mutation and careful pedigree segregation---but because the
paper's evidence shape is exactly the one the model was designed
for: a human-genetics variant-level claim with Mendelian
segregation and a clinical phenotype. The schema accommodates it
without strain. Taken together with the ten stress-test
candidates from Inouye, the six-paper pilot suggests that the
model covers conventional variant-level human-genetics evidence
comfortably but strains predictably at the population-genetics
and systems-biology boundaries of the field.

\subsection{Limitations}
\label{sec:disc-limitations}

The pilot has significant limitations, some inherent to the
design and some that a larger study could address.

First, six papers and a single annotator do not support claims
about inter-annotator agreement or generalisability. We treat
the work as a feasibility study
(\cref{sec:workflow-discussion}), not a benchmark. A larger
corpus annotated by multiple curators would be required to
benchmark the workflow against inter-annotator agreement
metrics.

Second, the model has not yet been released as a fully OWL-formalised
ontology. The present SHACL distribution covers top-level class
declarations, dimension enumerations, and the conditional
activation rules, but does not yet attempt the axiomatisation
required for classification and subsumption inference. A
LinkML source-of-truth representation, from which OWL, SHACL,
JSON Schema, and Avro could be generated, is one possible
direction for that work and would also make the model
interoperable with the existing OBO tool chain.

Third, our evaluation is not a clinical deployment. The model is
motivated by variant interpretation in clinical-genomics
programs (the BGM/UDN setting cited in the curation provenance
blocks), but the current pilot stops short of applying the model
to a live interpretation workflow. Whether annotations in this
form are useful to a working ACMG-AMP-style variant
classification pipeline remains a question for future work.

Fourth, the corpus is skewed toward research papers about
disease genes and variants. We did not annotate clinical-trial
evidence, family-reported phenotype data, or gene-dosage findings,
all of which would stress different parts of the schema. The
candidate extensions in \cref{tab:candidates} should be read as
findings from \emph{this} corpus, not as a complete survey of the
schema's coverage gaps.

\subsection{Future Work}
\label{sec:disc-future}

We see three near-term directions.

First, the promotion rule of \cref{sec:workflow-promotion} is
currently conservative: a candidate extension promotes to the
schema only after independent confirmation in a second paper. The
next natural step is to apply the rule forward by annotating a
second corpus selected to exercise the strongest candidates: a
handful of additional polygenic-score papers would settle CE-IN1
through CE-IN6, a handful of additional protein-interaction
papers would settle CE-J1 through CE-J3, and a handful of
additional GWAS follow-ups would settle CE-DU1.

Second, a LinkML or OWL source-of-truth would open the schema to
the OBO and Bridge2AI ecosystems and would let us align our
dimensions formally to existing vocabularies (ECO, SEPIO, HPO,
IAO, OBI, PROV-O). The detailed mapping is work that would
benefit from a collaboration with the relevant ontology
maintainers rather than being done in isolation.

Third, a clinical-deployment study---applying the model to a live
variant interpretation workflow---would turn the feasibility
claim into a utility claim. The specific question is whether
annotations in this format make a measurable difference to
clinical reasoning: for example, whether an interpreter working
from a pre-annotated set of publications reaches an ACMG-AMP
classification faster, more consistently, or with fewer
omissions than one working from the raw publications alone. The
expected effect size is modest, and the study would need
careful design, but the pilot annotations are a reasonable
starting point for a pre-registered comparison.
