% =====================================================================
%  SN2 (Supplementary Notes): the annotation and review protocol suite.
%  Fragment \input by supplement.tex under the Supplementary Notes part.
%  No preamble here.
% =====================================================================
\section{The annotation and review protocol suite}
\label{sec:supp-protocol-suite}

The annotation workflow described in the main paper has been
consolidated into a versioned, openly published protocol suite,
available in the project
repository.\footnote{\url{https://github.com/ForomePlatform/genetic-evidence-model/tree/master/protocols}}
The suite separates \emph{what} a well-formed annotation or review must
contain from \emph{how} it is executed (manually or with AI assistance),
and is used in both tracks of the pilot: the manual curator-authored
track and the AI-drafted track.

A mode-agnostic annotation protocol (\texttt{PROTOCOL.md}) states the
requirements for a valid annotation (classes, dimensions, source
anchoring, and flag semantics). Two mode-specific specialisations adapt
it for autonomous drafting (\texttt{PROTOCOL\_AUTONOMOUS.md}) and for
interactive, curator-in-the-loop work. A separate review protocol
(\texttt{REVIEW\_PROTOCOL.md}, with an interactive variant) governs
assertion-by-assertion review of drafted annotations. A document of
labelling examples records worked cases for the recurring judgement
calls. These written protocols are also packaged as executable skills
(defined in \cref{sec:supp-skills}), one for annotation and one for
review, so that the same rules can be applied by a human curator or by
an AI assistant working under curator supervision.

\subsection{Manual-track protocol}

In the \emph{manual track}, four of the six publications (Jossin 2017,
Davis 2011, Nelson 1992, Gupta 2015) were annotated by Vladimir
Seplyarskiy, a researcher in human and population genetics specialising
in human germline mutation and mutational processes (Assistant
Professor at UT~Southwestern; formerly Harvard Medical School and
Brigham and Women's Hospital). This expertise covers the
human-genetics and population-genetics knowledge domains exercised by
the corpus.

For each paper, the curator:

\begin{enumerate}
  \item read the PDF in full and marked relevant passages as highlights
        and short callouts;
  \item exported the marks with a helper script that converts PDF
        annotations into a structured JSON record (page, offset,
        highlighted text, free-text callout);
  \item constructed a summary table with one row per paper capturing
        high-level judgements (knowledge domain, method, credibility,
        resolution, mode of inheritance, and so on); and
  \item mapped the highlights and summary row into the schema,
        producing a YAML file conforming to the
        \texttt{GeneticEvidence} model.
\end{enumerate}

Whenever this mapping required a normalisation step, for example
turning the free-text ``localisation and protein expression'' into the
enumeration values \texttt{LOCALIZATION} and \texttt{EXISTENCE}, the
change was recorded inline as a \texttt{normalization\_note} with
\texttt{type: ai\_normalization} and the original text preserved.
These notes make later audit and ontology alignment straightforward.

All manual-track annotations were then validated against the SHACL
schema (\fpath{schema/genetic_evidence.shacl.ttl}). Violations, such
as a missing always-required dimension or a conditional dimension whose
activation condition fired but whose value was absent, had to be
resolved before the annotation was admitted to the corpus.

The four manual annotations are treated as \emph{ground truth} for the
purposes of this work. They are not a gold standard in the statistical
sense, because there is still only one annotator, but they are the
authoritative reference against which the AI-drafted track is
evaluated.

\subsection{AI-drafted-track protocol}

In the \emph{AI-drafted track}, two publications (Duerr 2006 and
Inouye 2018) were annotated directly from their PDFs by an AI assistant
following the same schema, using the four manual annotations as worked
examples.

Concretely, the assistant is provided with:

\begin{itemize}
  \item the schema (\fpath{schema/dimensions.md} and the SHACL file),
  \item the four manual annotations,
  \item the source PDFs, and
  \item the autonomous-annotation skill (see \cref{sec:supp-skills}).
\end{itemize}

It then produces draft YAML annotations in the same format as the
manual ones. When the assistant is uncertain about a mapping or has to
make a judgement call that a human curator might reasonably contest, it
records this inline using two flag types:

\begin{itemize}
  \item \texttt{ai\_uncertainty} for cases where the mapping is judged
        unreliable and a human decision is preferred; and
  \item \texttt{ai\_assumption} for judgements made without explicit
        support in the text that should be revisited at review time
        rather than silently absorbed.
\end{itemize}

The AI-drafted annotations are \emph{not} accepted as is. Each was
reviewed assertion by assertion by one of the authors, a curator with
expertise in genomic data curation and the evidence model rather than
in human or population genetics, operating under the review skill. The
reviewer checked:

\begin{itemize}
  \item faithful representation of the source text;
  \item conformance to the schema, including conditional-activation
        rules; and
  \item appropriateness of any \texttt{ai\_uncertainty} or
        \texttt{ai\_assumption} flags.
\end{itemize}

Corrections were applied directly to the YAML, with every change and
ruling logged under \fpath{annotations/reviews/}. Table~ST1
summarises the resulting verdicts per \texttt{GeneticEvidence} item,
and Supplementary Note~SN4 analyses them qualitatively. In all cases,
the \emph{curator}, not the model, is the authority for the recorded
annotations.

\subsection{Source anchoring and validation}

A central epistemic commitment of the workflow is that every
\emph{assertion} must be independently auditable back to its source.

\begin{itemize}
  \item Each assertion in a \texttt{GeneticEvidence} item carries a
        \texttt{source\_span} with a page number and a concise key
        phrase (typically under fifteen words) sufficient to locate the
        passage by search in the PDF.
  \item The phrase length limit is chosen to respect fair-use norms for
        copyrighted publisher material while still making verification
        practical.
  \item The SHACL schema treats \texttt{source\_span} as mandatory, so
        an annotation with any assertion lacking a source anchor is
        rejected at validation time.
\end{itemize}

Across the six-paper pilot, 95 of 95 assertions (100\%) carry source
spans by construction. Where a verbatim key phrase could not be
reconstructed from the public PDF, for example a value taken from a
table in a publisher PDF that is not checked into the repository, the
annotation records this explicitly via a
\texttt{source\_span\_is\_paraphrase} flag and an explanatory note, so
that future curators can refine the anchor against the original file.

\subsection{Provenance and protocol versions}

Two provenance points are important for interpreting the corpus.

First, the pilot corpus described in the main paper was initially
annotated under an earlier procedure (``protocol~v0''). The suite
described here is the subsequent, generalised formalisation. The pilot
annotations remain valid under the current protocol, but some artefacts,
such as slightly different span-selection conventions in the earliest
AI-drafted annotations, are recorded as forward-looking notes rather
than retro-fit edits.

Second, the AI-drafted annotations were drafted with Claude
Opus~4.7 (Duerr) and~4.8 (Inouye) and reviewed with Claude Opus~4.8,
both operating under the executable skills. These implementation details are recorded only for
reproducibility; they are not central to the scientific contribution.
In all cases, every AI-produced annotation and every review verdict was
adjudicated by the human curator, who remains the authority for the
recorded annotations.
