% =====================================================================
%  SN4 (Supplementary Notes): AI-track review outcomes.
%  Referenced from the main text (sec:workflow-flags) as Supplementary Note SN4.
% =====================================================================
\section{AI-track review outcomes}
\label{sec:supp-ai-review}

This note records the outcome of the independent expert review of the
two AI-drafted annotations, Duerr et al.\ 2006 (the clean GWAS exemplar)
and Inouye et al.\ 2018 (the polygenic-score stress test). Each draft was
reviewed assertion by assertion, in interactive mode, against
REVIEW\_PROTOCOL v1.0. The authoritative logs, with every flag and
ruling, are in the repository under \texttt{annotations/reviews/}; the
table below summarises the verdicts over the drafted items.

\begin{table}[h]
  \centering
  \small
  \begin{tabular}{lcccccc}
    \toprule
    Paper & Drafted & Approved & With flags & Edited & Rejected & Added on review \\
    \midrule
    Duerr 2006   & 6 & 0 & 2 & 4 & 0 & 1 (approved) \\
    Inouye 2018  & 3 & 1 & 1 & 1 & 0 & 1 (edited) \\
    \midrule
    Total        & 9 & 1 & 3 & 5 & 0 & 2 \\
    \bottomrule
  \end{tabular}
  \caption[Per-paper AI-track review verdicts]{Per-paper review verdicts over the AI-drafted items. ``With
  flags'' is approved-despite-non-blocking-flags (no edits applied);
  ``Added on review'' is evidence the reviewer found the draft had
  omitted and entered as a new item.}
  \label{tab:ai-review}
\end{table}

The Duerr row reflects the \texttt{v0} annotation; the Inouye row, and
all Inouye-related analyses here, use the \texttt{v1} re-annotation
produced under the protocol matured on the Duerr review (the original
five-item \texttt{v0} Inouye draft is retained in the repository for
comparison).

The two drafts were produced under different protocol versions, and the
later draft required markedly less correction. Two of Inouye's three
items were accepted without edits and one was edited, against four of
Duerr's six edited. Because Inouye is also the harder, polygenic-score
case, the comparison is confounded by paper as well as by protocol and
is offered as an observation rather than a controlled result; it
nonetheless agrees with the provenance signal, where Inouye's
thirty-seven cited spans were all verbatim and within the length limit,
while Duerr's earlier-protocol spans carried notation drift and a single
wrong page citation (recorded as forward notes, since they predate the
current rules).

A second pattern cuts across both papers. The items the reviewer had to
add were not granular association findings, which the drafts captured
well, but higher-level evidence: in Inouye, the paper's headline thesis
that combining component scores outperforms each alone; in Duerr, the
cited anti-p40 antibody trial that the discussion leans on for its
therapeutic-target argument. The drafts were thus most reliable on
specific, directly stated claims and weaker on a paper's synthetic
conclusions and cited translational evidence, the cross-item and
interpretive judgement that the human reviewer supplied.

The edits to existing items were concentrated in schema-level judgement,
not in the factual content of the assertions: knowledge-domain
assignment (family-based versus allele-frequency evidence), method
placement (a targeted replication recorded as an association study
rather than a fresh scan), credibility calibration (the score-combination
item lowered from high to medium for want of a stated mechanism), and the
applicability of variant ascertainment to a polygenic-score target. In
the last case the reviewer set the dimension to not applicable and
recorded the resulting constraint as support for a candidate extension.
No item was rejected, and no item was found to misrepresent its source.

One consequence follows for the promotion record. With variant
ascertainment no longer exercised by Inouye, its promotion rests on its
first proposal in Davis and its independent reuse in Nelson and Duerr,
which still satisfies the two-paper rule. Applying any change to the
promotion record is a separate curation step and is not made here. Both
reviews recommend a second pass over the changed items, since each paper
gained a new item and several existing items were edited.
