% =====================================================================
%  Supplementary Notes (SN6): per-paper case reports for the two
%  AI-drafted, curator-reviewed annotations (Duerr 2006, Inouye 2018).
%  Referenced from the main text (sec:examples) as Supplementary Note SN6.
%
%  This note gives the item-level detail the main paper omits for space:
%  a decomposition table, one worked annotation item, and the candidate
%  extensions each paper surfaces. Review verdicts are in SN4 and the
%  term-level crosswalk in SN5; the complete YAML annotations are in
%  Appendix SA2 (SA2.5 Duerr, SA2.6 Inouye). The YAML excerpts below are
%  abridged for readability; the verbatim files are in those appendices.
% =====================================================================
\section{Case reports: Duerr 2006 and Inouye 2018}
\label{sec:supp-crep}

The two AI-drafted annotations bracket the model's operating range, and
the main paper (Section~4) summarises them; this note gives the
item-level detail omitted there for space. Duerr is the clean case: six
of its seven \GeneticEvidence{} items map without strain, and the one
genuine gap is translational. Inouye is the stress test: every item
stretches the variant-level model, which is why it surfaces six
candidate extensions. The assertion-by-assertion review verdicts for
both are in \cref{sec:supp-ai-review}, the term-level standards mapping
is in \cref{sec:supp-crosswalk}, and the complete annotations are in
Appendix~SA2 (SA2.5 Duerr, SA2.6 Inouye).

\subsection{Case Report CR-DU: Duerr et al.\ 2006 (genome-wide association)}
\label{sec:supp-ex-duerr}

Duerr and colleagues identified \textit{IL23R} as an inflammatory bowel
disease gene by a genome-wide association scan in non-Jewish ileal
Crohn's disease, with replication in a Jewish cohort and family-based
transmission testing in 883 nuclear families. Our annotation decomposes
the paper into seven \GeneticEvidence{} items (\cref{tab:cr-duerr}).

\begin{table}[H]
  \centering
  \footnotesize
  \caption[Duerr 2006: evidence-item decomposition]{Duerr 2006: the
  seven \GeneticEvidence{} items, with target type, method (leaf-level
  per the hierarchy), and credibility.}
  \label{tab:cr-duerr}
  \begin{tabularx}{\linewidth}{@{}l >{\raggedright\arraybackslash}X >{\raggedright\arraybackslash}p{1.35cm} >{\raggedright\arraybackslash}p{2.5cm} l@{}}
    \toprule
    ID & Claim & Target & Method & Cred. \\
    \midrule
    GE-1 & rs11209026 protective in non-Jewish ileal CD (discovery) & \dimval{Variant} & \dimval{GWAS} & High \\
    GE-2 & rs11209026 protective in Jewish ileal CD (replication) & \dimval{Variant} & \dimval{Association Study} & High \\
    GE-3 & Gln undertransmitted to affected offspring (FBAT) & \dimval{Variant} & \dimval{Transmission Disequilibrium Test} & High \\
    GE-4 & Multiple independent IL23R signals (conditional) & \dimval{Variant} & \dimval{Fine Mapping} & High \\
    GE-5 & No association at IL12RB1/IL23A/IL12B & \dimval{Gene} & \dimval{GWAS} & High \\
    GE-6 & Cited mouse-model and functional support & \dimval{Gene} & \dimval{In Vivo}, \dimval{In Vitro} & High \\
    GE-new-1 & Cited anti-p40 (IL12B) antibody trial & \dimval{Gene} & \dimval{Clinical Evidence} & High \\
    \bottomrule
  \end{tabularx}
\end{table}

Six of the seven items fit the schema as it stands. The seventh,
GE-new-1 (a cited anti-p40 antibody trial added during the
missing-evidence check), is drug-response evidence with no fitting
genetic knowledge domain, the translational-scope gap discussed below.
The discovery item GE-1, shown abridged below, illustrates how the
dimensions, the credibility comment, and the source-anchored assertions
are recorded.

\begin{lstlisting}[language=yaml]
id: GE-1
label: "rs11209026 (Arg381Gln) in IL23R confers protection
         against ileal CD in non-Jewish discovery cohort"
knowledge_domain: [POPULATION_GENETICS]
method: [GWAS]
target_type: VARIANT
target: "rs11209026 (c.1142G>A, p.Arg381Gln) in IL23R"
resolution: VARIANT
variant_ascertainment: [OBSERVED_IN_CASES, OBSERVED_IN_CONTROLS]
phenotype_scale: CLINICAL
credibility: HIGH
credibility_comment: >
  Genome-wide significance after Bonferroni correction
  (corrected P = 1.56e-3); large effect (OR = 0.26);
  independent replication follows in GE-2.
candidate_extension_exercised: CE-DU1   # direction of effect
assertions:
  - id: GE-1.A1
    statement: "rs11209026 is genome-wide significantly
                associated with ileal Crohn's disease"
    source_span: {page: 1,
      key_phrase: "rs11209026 (P = 5.05e-9, corrected P = 1.56e-3)"}
  - id: GE-1.A2
    statement: "The Gln allele is protective (OR = 0.26)"
    source_span: {page: 2,
      key_phrase: "glutamine allele appears to protect against
                   development of CD"}
\end{lstlisting}

\paragraph{Candidate extensions.} Curator review surfaced three, all
provisional under the two-papers promotion rule of the main paper:
\begin{itemize}
  \item \textbf{CE-DU1, direction of effect:} a first-class dimension
  distinguishing protective from risk-conferring alleles (here OR~0.26),
  currently implicit in the sign of the odds ratio; the distinction
  matters for ACMG/AMP-style downstream use.
  \item \textbf{CE-DU3, cross-item synthesis:} a construct for a
  higher-order claim that references several items jointly, here the
  paper's recommendation that blocking the IL-23 pathway is a rational
  therapeutic strategy; exercised twice (the recommendation and
  GE-new-1), so it is also a translational-scope question.
  \item \textbf{CE-DU4, conditional-activation scope:} widen the
  \{\texttt{mode\_of\_inheritance}, \texttt{mendelian\_segregation},
  \texttt{exact\_variant}, \texttt{subdomain}\} activation from
  \dimval{Human Genetics}~$+$~\dimval{Gene} to also include
  \dimval{Variant} (as in GE-3's family-based transmission test), with a
  \dimval{Complex} mode-of-inheritance value and a family-based subdomain
  value.
\end{itemize}
A fourth candidate, CE-DU2 (a negated-assertion dimension), was
\emph{retracted} on review: an assertion is a predicate and already
expresses absence, so no dimension is needed. The retraction is the
corpus's clearest example of the conservative promotion rule blocking a
false positive.

\subsection{Case Report CR-IN: Inouye et al.\ 2018 (polygenic risk score)}
\label{sec:supp-ex-inouye}

Inouye and colleagues built and externally validated a meta-genomic risk
score (metaGRS) for coronary artery disease from 1{,}745{,}180 variants,
testing it in 482{,}629 UK Biobank participants. The target is a derived
whole-genome score and the measurements are epidemiological (hazard
ratio, C-index), so every item strains a variant-level model. The
annotation used here is the \texttt{v1} re-annotation produced under the
current protocol, the one matured on the Duerr review (the original
five-item \texttt{v0} draft is retained in the repository for
comparison). It decomposes the paper into four claim-coherent items
(\cref{tab:cr-inouye}), each representable only through deliberate
``least-wrong'' placeholder choices that are recorded as forced fits and
candidate extensions rather than hidden.

\begin{table}[H]
  \centering
  \footnotesize
  \caption[Inouye 2018: evidence-item decomposition]{Inouye 2018
  (\texttt{v1}): the four \GeneticEvidence{} items. Every item forces the
  same placeholder target type and resolution (\dimval{Variant}), flagged
  against CE-IN1 and CE-IN2; the true target is a whole-genome polygenic
  score.}
  \label{tab:cr-inouye}
  \begin{tabularx}{\linewidth}{@{}l >{\raggedright\arraybackslash}X p{1.5cm} l@{}}
    \toprule
    ID & Claim & Target$^{\dagger}$ & Cred. \\
    \midrule
    GE-1     & Prediction and quintile stratification (HR 1.71/SD; 4.17-fold) & \dimval{Variant} & High \\
    GE-2     & Independence from six conventional risk factors & \dimval{Variant} & High \\
    GE-3     & Stratification persists among medicated individuals & \dimval{Variant} & High \\
    GE-new-1 & Combining three component scores outperforms each & \dimval{Variant} & Medium \\
    \bottomrule
  \end{tabularx}

  {\footnotesize $^{\dagger}$Placeholder; the true target is a polygenic
  \dimval{Score} (CE-IN1) at whole-genome-aggregate resolution (CE-IN2).
  Method is recorded at the intermediate \dimval{Association Study} /
  \dimval{Meta-Analysis} level (no leaf fits polygenic-score association;
  flag F-METH), and variant ascertainment is marked not-applicable (the
  CE-IN1 SHACL point).}
\end{table}

The independence item GE-2, shown abridged below, makes the strain
concrete: \texttt{target\_type} and \texttt{target\_resolution} are
forced to \dimval{Variant} (flagged against CE-IN1 and CE-IN2), variant
ascertainment is marked not-applicable (the CE-IN1 SHACL point), and the
epidemiological measures are carried as candidate-extension targets
(CE-IN4).

\begin{lstlisting}[language=yaml]
id: GE-2
label: "metaGRS predicts incident CAD largely independently
        of six conventional risk factors"
knowledge_domain: [POPULATION_GENETICS]
method: [ASSOCIATION_STUDY]    # intermediate; no leaf fits PGS assoc. (F-METH)
target_type: VARIANT           # forced; CE-IN1 proposes SCORE
target_resolution: VARIANT     # forced; CE-IN2 -> WHOLE_GENOME_AGGREGATE
variant_ascertainment: not_applicable_or_omitted  # CE-IN1 (SHACL); F-VA
phenotype_scale: CLINICAL
credibility: HIGH
# epidemiological measures (C-index, HR) in special_considerations; CE-IN4
assertions:
  - id: GE-2.A1
    statement: "metaGRS has a higher C-index than any single factor"
    source_span: {page: 6,
      phrase: "higher C-index (C = 0.623; 95% CI: 0.615 to 0.630)"}
  - id: GE-2.A3
    statement: "Adding metaGRS to all six factors raised C-index to 0.696"
    source_span: {page: 6,
      phrase: "C-index of 0.696 (95% CI: 0.688 to 0.703)"}
\end{lstlisting}

\paragraph{Candidate extensions.} The annotation surfaces six, the most
of any paper in the corpus:
\begin{itemize}
  \item \textbf{CE-IN1, \dimval{Score} target type:} a metaGRS is
  neither gene nor variant but a derived composite predictor; add a
  \dimval{Score} value (possibly subtyped polygenic vs composite). The
  v1 review sharpened this: under the forced \dimval{Variant} target the
  variant-ascertainment conditional is unsatisfiable either way (omitting
  it fires the rule; marking it not-applicable violates the shape), so
  only a \dimval{Score} target dissolves the contradiction.
  \item \textbf{CE-IN2, \dimval{Whole Genome Aggregate} resolution:} the
  score spans the whole genome non-contiguously, which the
  contiguous/pointwise resolution values do not capture.
  \item \textbf{CE-IN3, target-composition dimension:} orthogonal to
  resolution, a \dimval{Single}/\dimval{Aggregate}/\dimval{Composite}
  axis; a complementary framing of the same gap as CE-IN2.
  \item \textbf{CE-IN4, epidemiological measurement targets:} extend
  \texttt{measurement\_target} (today molecular-assay-oriented) with
  \dimval{Hazard Ratio}, \dimval{Odds Ratio}, \dimval{C-Index},
  \dimval{AUROC}, and the like, activated for population-genetics
  evidence.
  \item \textbf{CE-IN5, structured cohort descriptor:} replace free-text
  cohort properties with named fields (size, ancestry composition,
  follow-up, ascertainment); reusable across Duerr, Davis, and future
  cohort studies.
  \item \textbf{CE-IN6, natural vs derived-artefact target:} a flag (or
  \dimval{Natural}/\dimval{Derived} dimension) marking that the metaGRS
  is a human construction rather than a natural kind.
\end{itemize}
